Value Ordering, Discrepancies, and Restarts for Subgraph Algorithms

### 2018-04-20 paper/paper.tex
Tweaks

The subgraph isomorphism problem is to find a copy of a small ``pattern'' graph inside a larger
``target'' graph.  The current single strongest subgraph isomorphism solver uses ``highest degree
first'' as a value-ordering heuristic to direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and restarts (which are
extremely beneficial). To explore different parts of the search space when restarting, we introduce
a new value-ordering heuristic, which adds a small amount of randomness to the ``highest degree
first'' heuristic.  The effects of these changes are shown in
\figureref{figure:scatter-by-family}: we can make a state-of-the-art algorithm perform much better on a
large number of satisfiable instances, whilst performing worse only rarely on satisfiable instances,
and never on unsatisfiable instances.  Further, these properties are \emph{not} shared by
discrepancy searches (\figureref{figure:scatter-dds}) or by random value-ordering with restarts
(\figureref{figure:scatter-random}). Finally, we show that this strategy is also effective in an
optimisation setting, producing benefits in two maximum common induced subgraph algorithms.

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \algorithmref{algorithm:sip}, we use
smallest domain first, tiebreaking on highest degree. Variable ordering is not the focus of this
paper---as far as possible, we use this same dynamic ordering throughout the paper, so as not to
affect the performance of the algorithm on unsatisfiable instances.  For value ordering, which
occurs on \siplineref{valueordering} of \algorithmref{algorithm:sip}, we will initially select vertices from
highest degree to lowest degree; we return to value ordering in the second half of this paper. Note
that the variable-ordering heuristic affects both unsatisfiable and satisfiable instances, whilst
the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{\url{https://github.com/ciaranm/ijcai2018-restarting-sip-paper}, DOI to be issued pre-publication}
\algorithmref{algorithm:sip} in C++ using the bit-parallel implementation of
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

@@

We begin by verifying that our implementation of the basic \algorithmref{algorithm:sip} is competitive. In
\figureref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \algorithmref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\cite{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout; the horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

@@

    \caption{The cumulative number of satisfiable instances solved over time, using four
    different value-ordering heuristics. The degree and biased lines are close together.}
    \label{figure:value-ordering-heuristics}
\end{figure}

@@

\subsection{Adding Restarts to the Algorithm}

@@

We use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used SMAC \cite{DBLP:conf/lion/HutterHL11} to select the value 660. (We also tried a geometric
restart scheme, which gave less favourable results.)

The remainder of the changes are to introduce nogoods. To avoid exploring portions of the search
space that we have already visitied, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees. We generate simple decision nogoods, as follows. The $B$
argument to $\FuncSty{search}$ tracks the branching choices made so far, in the form of a CNF clause
with each literal being a guessed $(v \mapsto v')$ assignment. Upon backtracking due to a decision to
restart, we post a nogood of the form $(v \mapsto v') \wedge (w \mapsto w') \wedge (x \mapsto x')
\Rightarrow \bot$ for every branch to the left of the current (incomplete) branch in the search tree.

@@

With McSplit${\downarrow}$, the inability to use two watched literals means that in many cases we
introduce a small slowdown. However, the overall pattern is the same: when introducing restarts and
a biased value ordering heuristic, it is much more common to see a large speedup than it is a large
slowdown.

@@

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances.

The success of this approach raises a number of interesting questions and avenues for possible
future research. For example, will this same technique work for other problems where we have a good
value-ordering heuristic? Are there further benefits to be had from also modifying variable-ordering
heuristics? Are there theoretical rather than just empirical reasons for using the
\emph{softmax}-like bias scheme? And how can such a search process be parallelised?

>>>>

The subgraph isomorphism problem is to decide whether a copy of a small ``pattern'' graph inside a
larger ``target'' graph. Although \NP-complete, by combining design techniques from artificial
intelligence with careful algorithm engineering, modern subgraph isomorphism solvers can often
produce exact solutions quickly even on graphs with thousands of vertices. The current single
strongest subgraph isomorphism solver uses ``highest degree first'' as a value-ordering heuristic to
direct a constraint programming style search
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16,DBLP:conf/ijcai/McCreeshPT16}. This
heuristic is much better than branching randomly, but is still far from perfect.  We could try to
introduce tiebreaking mechanisms and similar refinements to strengthen the heuristic. However, given
the extreme cost of making an incorrect branching choice early in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search (which performs poorly), and a new form of search
(which is extremely beneficial) based around restarts and nogood recording. To explore different
parts of the search space when restarting, we introduce a novel value-ordering heuristic, which adds
a small amount of randomness to the ``highest degree first'' heuristic.  The effects of these
changes are shown in \figureref{figure:scatter-by-family}: we can make a state-of-the-art algorithm
perform much better on a large number of satisfiable instances, whilst performing worse only rarely
on satisfiable instances, and never on unsatisfiable instances.  Further, these properties are
\emph{not} shared by discrepancy searches (\figureref{figure:scatter-dds}) or by random
value-ordering with restarts (\figureref{figure:scatter-random}). Finally, we show that this
strategy is also effective in an optimisation setting, producing benefits in two maximum common
induced subgraph algorithms.

@@

For the variable-ordering heuristic on \siplineref{variableordering} of
\algorithmref{algorithm:sip}, we use smallest domain first, tiebreaking on highest degree
\cite{DBLP:journals/ai/HaralickE80,DBLP:journals/jair/McCreeshPST18}. Variable ordering is not the
focus of this paper---as far as possible, we use this same dynamic ordering throughout the paper, so
as not to affect the performance of the algorithm on unsatisfiable instances.  For value ordering,
which occurs on \siplineref{valueordering} of \algorithmref{algorithm:sip}, we will initially select
vertices from highest degree to lowest degree; we return to value ordering in the second half of
this paper. Note that the variable-ordering heuristic affects both unsatisfiable and satisfiable
instances, whilst the value-ordering heuristic is relevant only for satisfiable instances.

@@

Our experiments are performed on systems with dual Intel Xeon E5-2697A v4 CPUs and 512GBytes RAM,
running Ubuntu 17.04. We
implemented\footnote{\url{https://github.com/ciaranm/restarting-sip-paper}, DOI to be issued pre-publication}
\algorithmref{algorithm:sip} in C++ using the bit-parallel implementation of
Hoffmann et al.\ \cite{DBLP:conf/aaai/HoffmannMR17} as a starting point, and compiled it using GCC 6.3.0. We use a
deterministic pseudo-random number generator for reproducibility.

@@

We begin by verifying that our implementation of the basic \algorithmref{algorithm:sip} is competitive. In
\figureref{figure:others} we show the cumulative number of instances solved over time, using our
implementation (both with the degree heuristic, and with the modifications described in the
remainder of this paper), the Glasgow2 and Glasgow3 algorithms from which \algorithmref{algorithm:sip} is derived
\cite{DBLP:conf/cp/McCreeshP15,DBLP:conf/lion/KotthoffMS16}, the PathLAD variation of the LAD
algorithm \cite{DBLP:journals/ai/Solnon10,DBLP:conf/lion/KotthoffMS16}, and VF2
\cite{DBLP:journals/pami/CordellaFSV04}, in each case using the original implementation provided
by the algorithm's authors. For each algorithm, the $y$ value
gives the cumulative number of instances which (individually) can be solved in no more than $x$
milliseconds.  The vertical distance between two lines therefore shows how many more instances can
be solved by one solver than another, if every instance is run separately with the chosen $x$
timeout. The horizontal distance shows how many times longer the per-instance timeout would need to
be to allow the rightmost algorithm to succeed on $y$ out of the 14,621 instances (bearing in mind
that the two sets of $y$ instances could be different).

@@

    \caption{The cumulative number of satisfiable instances solved over time, using four
    different value-ordering heuristics.}
    \label{figure:value-ordering-heuristics}
\end{figure}

@@

\subsection{Adding Restarts and Nogood Recording to the Algorithm}

@@

We use the Luby scheme \cite{DBLP:journals/ipl/LubySZ93} to determine when to restart. We count the
number of backtracks (that is, when we reach the end of the main $\KwSty{foreach}$ loop on
\siplineref{countrestarts}) to decide when to restart, rather than counting recursive calls.
Following established wisdom, we multiply each item in the Luby sequence by a magic constant---we
used the SMAC automatic parameter tuner \cite{DBLP:conf/lion/HutterHL11} to select the value 660.
(We also tried a geometric restart scheme, which gave less favourable results.)

The remainder of the changes are to introduce nogoods. To avoid exploring portions of the search
space that we have already visitied, every time we restart, we add new constraints to the problem
which eliminate already-explored subtrees---such a constraint is called a nogood. We generate simple
decision nogoods, as follows. The $B$ argument to $\FuncSty{search}$ tracks the branching choices
made so far, in the form of a CNF clause with each literal being a guessed $(v \mapsto v')$
assignment. Upon backtracking due to a decision to restart, we post a nogood of the form $(v \mapsto
v') \wedge (w \mapsto w') \wedge (x \mapsto x') \Rightarrow \bot$ for every branch to the left of
the current (incomplete) branch in the search tree, and when we first make a decision to restart
before backtracking, we post a similar nogood eliminating the entire subtree explored.

@@

With McSplit${\downarrow}$, the inability to use two watched literals means that in many cases we
introduce a small slowdown. However, the overall pattern is the same: when introducing restarts and
a biased value ordering heuristic, it is much more common to see a large speedup than a large
slowdown.

@@

We have shown that it is both possible and beneficial to introduce a small amount of randomness into
backtracking search algorithms for hard subgraph problems, without compromising existing search
order heuristics, and without paying a penalty for unsatisfiable instances. In principle it is
well-known that introducing some form of randomness \cite{DBLP:conf/aaai/GomesSK98} or reduced
commitment \cite{DBLP:conf/ijcai/HarveyG95} into backtracking search can be beneficial, but until
now it has not been clear how to do so efficiently in practice, particularly when we already have
strong search ordering heuristics and when are expecting to encounter large numbers of unsatisfiable
instances. Our results show that the key to success is a combination of aggressive restarts, nogood
recording, and biasing an existing value-ordering heuristic with just a small amount of randomness.

The effectiveness of this approach raises a number of interesting questions and avenues for possible
future research. For example, will this same technique work for other problems where we have a good
value-ordering heuristic? Are there further benefits to be had from also modifying variable-ordering
heuristics? Are there theoretical rather than just empirical reasons for using the
\emph{softmax}-like bias scheme? And how can such a search process be parallelised?
